---
title: "A Quick Introduction to PhD via Examples"
author: "Anne Chao and Chun-Huo Chiu"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{A Quick Introduction to PhD via Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning=FALSE,}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, message = FALSE,
  comment = ""
)
library(PhD)
# devtools::load_all(".")
library(ggplot2)
#library(kableExtra)
library(knitr)
data(data.abu)
data(data.inc)
```
`PhD` (**Ph**ylogenetic **D**iversity) is a R package providing the rarefaction/extrapolation and asymptotic estimation to achieve fair comparison of abundance-sensitive phylogenetic diversity among multiple assemblages (Hsieh and Chao, Systematic Biology, 2017). In this document, we provide a quick introduction demonstrating how to run `PhD`. The Detailed information about `PhD` functions is described in the `PhD` Manual, also available in [CRAN](https://cran.r-project.org/package=PhD). An online version of `PhDOnline` (<https://chao.shinyapps.io/PhDOnline/>) is developed for users without R background. The theoretical derivation of phylogenetic diversity measurements was proposed in Chao, A., Chiu C.-H. and Jost, L. (2010). `PhD` is an extension for [`iNextPD`](https://cran.r-project.org/package=iNextPD), which computes diversity estimates and its `conf` confidence intervals for the two types of rarefaction and extrapolation (R/E) based on two different sampling data sets, individual-based abundance data and quadrat-based incidence data.

Besides `iNEXTPD`, there are three main functions in `PhD`, including `PhdAsy`, `PhdObs` and `estimatePD`.

1. `iNEXTPD`: the phylogenetic extension of `iNEXT`, focusing on the seamless rarefaction and extrapolation sampling curves of phylogenetic diversity(PD) or phylogenetic Hill number(D) with respect to default/specified diversity order `q` <font size="1">$\ge$</font> 0 and reference time `reftime`. See Chao et al. (2010, 2015) and Hsieh and Chao (2017) for pertinent background and methods.

2. `PhdAsy`: estimates asymptotic diversity profile with respect to specified/default order `q` and reference time `reftime` to infer true phylogenetic diversity(PD) or phylogenetic Hill number(D). It is based on statistical estimation of the unknown true phylogenetic diversity; see Chao et al. (2015) and Hsieh and Chao (2017) for the statistical estimation detail.

3. `PhdObs`: computes emprircal phylogenetic diversity(PD) or phylogenetic Hill number(D) for specified/default order `q` and reference time `reftime`.

4. `estimatePD`: the phylogenetic extension of `estimateD` in `iNEXT` package, aims to estimate phylogenetic diversity(PD) or phylogenetic Hill number(D) with particular user-specified levels of sample coverage with respect to specified/default order `q` and reference time `reftime`.

<br><font size="5">**SOFTWARE NEEDED TO RUN PhD IN R:**</font>
--------------------------------------

- Required: [R](https://cran.r-project.org/)
- Suggested: [RStudio IDE](https://www.rstudio.com/products/RStudio/#Desktop)

<br><font size="5">**HOW TO RUN PhD:**</font>
-------------------

The `PhD` package is available on [CRAN](https://cran.r-project.org/package=PhD) and can be downloaded with a standard R installation procedure using the following commands. Before installation, the additional visualization and computation extension packages (`chaoUtility`,`ape`, `dplyr`, `ggplot2`, `phyclust`, `Rcpp`) must be loaded. Please note that currently, package `chaoUtility` can only be available via github instead of CRAN and thus the installation procedure slightly differs from the one of standard R installation.

```{r eval=FALSE}
## install chaoUtility package from github
install.packages('devtools')
library(devtools)
install_github('chaolab2019/chaoUtility')

## install PhD package from CRAN
install.packages("PhD")

## Or instead, install the latest version from github
install_github('AnneChao/PhD')

## import packages
library(chaoUtility)
library(PhD)
```
<br><font size="5"><b>DATA FORMAT/INFORMATION</b></font> 
-------------------
In PhD package, all four main functions share the same required format of `data`. Two types data are supported.

1. Individual‐based abundance data(`datatype="abundance"`): a S by N matrix/data.frame where N is the number of assemblages. The element in i-th row and k-th is the abundance of species i in assemblage k. Please note that the rownames of `data` must be the species names matching the species names in phylogeny tree and thus can't be empty. Users can use the following command to see an example data containing three assemblages.
```{r eval=FALSE}
# include abundance data along with its phylogeny tree
data(data.abu)
data.abu$data
```

2. Sampling‐unit‐based incidence raw data(`datatype="incidence_raw"`): the sampling unit is quadrat or transect, the observed species was only recorded as presence(detection)/absence(non-detection) data in each sampling unit. For single assemblage consisting of `t_` sampling units, `data` is a S by `t_` incidence matrix, where S is the number of species. When there are N assemblages, users must first merge N data matrices by species identity to obtain a large incidence matrix, where the rows of the matrix refer to all species presented in the pooled data. Likewise, the rownames of `data` must be the species names matching the species names in phylogeny tree and thus can't be empty. Users can use the following command to see an example data containing two assemblages.
```{r eval=FALSE}
# include incidence raw data along with its phylogeny tree
data(data.inc)
# an incidence matrix of 29 columns, where:
# the first assemblage consists of the first 12 columns (sampling units) and
# the second assemblage consists of the left 17 columns (sampling units).
data.inc$data
# the number of sampling units for two assemblages are 12 and 17, respectively.
data.inc$t
```

<br><font size="5"><b>MAIN FUNCTIONS:</b></font>
-------------------

We describe the four main functions with default arguments below :

1. <u><font size="3">**iNEXTPD()**</u></font>:

```{r, eval=FALSE}
iNEXTPD(data, t_, datatype = "abundance",tree,q = c(0,1,2),reftime=NULL,type = 'PD',
        endpoint = NULL, knots = 40, size = NULL, nboot = 50, conf = 0.95)
```
The arguments of this function are described below.

<table style="width:100%;">
<colgroup>
<col width="20%">
<col width="80%">
</colgroup>
<thead>
<tr>
<th align="center">Argument</th>
<th align="left">Description</th>
</tr>
</thead>
<tbody>
<tr>
<td align="center"><code>data</code></td>
<td align="left"> a `matrix`/`data.frame` of species abundances or incidence data. See the <b>DATA FORMAT/INFORMATION</b> section above for detailed required format.</td>
</tr>
<tr>
<td align="center"><code>t_</code></td>
<td align="left"> needed only when `datatype = "incidence_raw"`, a sequence of named nonnegative integers specifying the sampling units in each community. Will be ignored if `datatype = "abundance"`.</td>
</tr>
<tr>
<td align="center"><code>datatype</code></td>
<td align="left"> the data type of input data: individual-based abundance data (`datatype = "abundance"`), or sampling-unit-based incidence matrix (`datatype = "incidence_raw"`). Default is `abundance`.</td>
</tr>
<tr>
<td align="center"><code>tree</code></td>
<td align="left"> a `phylo` object describing the Newick phylogeny tree for all observed species in the pooled assemblage.</td>
</tr>
<tr>
<td align="center"><code>q</code></td>
<td align="left"> a nonnegative value or sequence specifying the diversity order. There are three measures: Faith's PD (`q = 0`), a simple transformation of phylogenetic entropy (`q = 1`) and a simple transformation of Rao's quadratic entropy (`q = 2`). Default is `c(0,1,2)`.</td>
</tr>
<tr>
<td align="center"><code>reftime</code></td>
<td align="left"> a positive value or sequence specifying the reference time for tree. If `NULL`, `reftime` = the tree depth of pooled assemblage. Default is `NULL`.</td>
</tr>
<tr>
<td align="center"><code>type</code></td>
<td align="left"> desired diversity type: `'PD'` for phylogenetic diversity and `'D'` for phylogenetic Hill number. See Chao et al. (2010) for details. Default is `'PD'`.</td>
</tr>
<tr>
<td align="center"><code>endpoints</code></td>
<td align="left"> a positive interger specifying the sample size that is the endpoint for R/E calculatione. If `NULL`, `endpoint` = double of the reference sample sizes. Will be ignored if `size` is given. Default is `NULL`.</td>
</tr>
<tr>
<td align="center"><code>knots</code></td>
<td align="left"> a positive integer specifying the number of knots between 1 and the `endpoint`. Default is 40.</td>
</tr>
<tr>
<td align="center"><code>size</code></td>
<td align="left"> a sequence of positive integers specifying the sample sizes for which `PD` or `D` estimates will be calculated. If `NULL`, then estimates will be calculated for those sample sizes determined by the specified/default `endpoint` and `knots`.</td>
</tr>
<tr>
<td align="center"><code>nboot</code></td>
<td align="left"> a positive integer specifying the number of bootstrap replications. Enter 0 to skip bootstrap; in this case, the caculation of standard errors and confidence intervals will be skipped. Default is 50.</td>
</tr>
<tr>
<td align="center"><code>conf</code></td>
<td align="left"> a positive number < 1 specifying the level of confidence interval. Default is 0.95.</td>
</tr>

</tbody>
</table>

<br>

The sequence of sample sizes for which PD/D R/E curve will be estimated is determined by `endpoint`, `knots` and `size`.
If `size` is specified by users, then PD/D will be estimated for these sizes plus the reference sample size(total species abundances for abundance data; total sampling units for incidence data). 
If `knots=K`, then the sequece would be K/2 evenly‐spaced knots (sample sizes) before reference sample size and K/2 from reference sample size to `endpoint`, where the `endpoint` is default to double reference sample size. By default, `knots=40`.
If `endpoint` specified by users is smaller than the reference sample size, `iNEXTPD` only computes rarefaction estimates for approximately `knots` evenly spaced knots before the `endpoint`.


2. <u><font size="3">**PhdAsy()**</u></font>:

```{r, eval=FALSE}
PhdAsy(data, t_, datatype = "abundance", tree, q = seq(0, 2, by = 0.25), 
       reftime = NULL, type = 'PD', nboot = 50, conf = 0.95)
```

The arguments of this function are described below.

<table style="width:100%;">
<colgroup>
<col width="20%">
<col width="80%">
</colgroup>
<thead>
<tr>
<th align="center">Argument</th>
<th align="left">Description</th>
</tr>
</thead>
<tbody>
<tr>
<td align="center"><code>data</code></td>
<td align="left"> a `matrix`/`data.frame` of species abundances or incidence data. See the description of function `iNEXTPD` for more details.</td>
</tr>
<tr>
<td align="center"><code>t_</code></td>
<td align="left"> needed only when `datatype = "incidence_raw"`, a sequence of named nonnegative integers specifying the sampling units in each community. Will be ignored if `datatype = "abundance"`.</td>
</tr>
<tr>
<td align="center"><code>datatype</code></td>
<td align="left"> the data type of input data: individual-based abundance data (`datatype = "abundance"`), or sampling-unit-based incidence matrix (`datatype = "incidence_raw"`). Default is `abundance`.</td>
</tr>
<tr>
<td align="center"><code>tree</code></td>
<td align="left"> a `phylo` object describing the Newick phylogeny tree for all observed species in the pooled assemblage.</td>
</tr>
<tr>
<td align="center"><code>q</code></td>
<td align="left"> a nonnegative value or sequence specifying the diversity order. Default is `seq(0, 2, by = 0.25)`.</td>
</tr>
<tr>
<td align="center"><code>reftime</code></td>
<td align="left"> a positive value or sequence specifying the reference time for tree. If `NULL`, `reftime` = the tree depth of pooled assemblage. Default is `NULL`.</td>
</tr>
<tr>
<td align="center"><code>type</code></td>
<td align="left"> desired diversity type: `'PD'` for phylogenetic diversity and `'D'` for phylogenetic Hill number. See Chao et al. (2010) for details. Default is `'PD'`.</td>
</tr>
<tr>
<td align="center"><code>nboot</code></td>
<td align="left"> a positive integer specifying the number of bootstrap replications. Enter 0 to skip bootstrap; in this case, the caculation of standard errors and confidence intervals will be skipped. Default is 50.</td>
</tr>
<tr>
<td align="center"><code>conf</code></td>
<td align="left"> a positive number < 1 specifying the level of confidence interval. Default is 0.95.</td>
</tr>

</tbody>
</table>

<br>

3. <u><font size="3">**PhdObs()**</u></font>:

```{r, eval=FALSE}
PhdObs(data, t_, datatype = "abundance", tree, q = seq(0, 2, by = 0.25),
       reftime = NULL, type = "PD",nboot = 50,conf = 0.95)
```

The arguments of this function are described below.

<table style="width:100%;">
<colgroup>
<col width="20%">
<col width="80%">
</colgroup>
<thead>
<tr>
<th align="center">Argument</th>
<th align="left">Description</th>
</tr>
</thead>
<tbody>
<tr>
<td align="center"><code>data</code></td>
<td align="left"> a `matrix`/`data.frame` of species abundances or incidence data. See the description of function `iNEXTPD` for more details.</td>
</tr>
<tr>
<td align="center"><code>t_</code></td>
<td align="left"> needed only when `datatype = "incidence_raw"`, a sequence of named nonnegative integers specifying the sampling units in each community. Will be ignored if `datatype = "abundance"`.</td>
</tr>
<tr>
<td align="center"><code>datatype</code></td>
<td align="left"> the data type of input data: individual-based abundance data (`datatype = "abundance"`), or sampling-unit-based incidence matrix (`datatype = "incidence_raw"`). Default is `abundance`.</td>
</tr>
<tr>
<td align="center"><code>tree</code></td>
<td align="left"> a `phylo` object describing the Newick phylogeny tree for all observed species in the pooled assemblage.</td>
</tr>
<tr>
<td align="center"><code>q</code></td>
<td align="left"> a nonnegative value or sequence specifying the diversity order. Default is `seq(0, 2, by = 0.25)`.</td>
</tr>
<tr>
<td align="center"><code>reftime</code></td>
<td align="left"> a positive value or sequence specifying the reference time for tree. If `NULL`, `reftime` = the tree depth of pooled assemblage. Default is `NULL`.</td>
</tr>
<tr>
<td align="center"><code>type</code></td>
<td align="left"> desired diversity type: `'PD'` for phylogenetic diversity and `'D'` for phylogenetic Hill number. See Chao et al. (2010) for details. Default is `'PD'`.</td>
</tr>
<tr>
<td align="center"><code>nboot</code></td>
<td align="left"> a positive integer specifying the number of bootstrap replications. Enter 0 to skip bootstrap; in this case, the caculation of standard errors and confidence intervals will be skipped. Default is 50.</td>
</tr>
<tr>
<td align="center"><code>conf</code></td>
<td align="left"> a positive number < 1 specifying the level of confidence interval. Default is 0.95.</td>
</tr>


</tbody>
</table>

4. <u><font size="3">**estimatePD()**</u></font>:

```{r, eval=FALSE}
estimatePD <- function(data,t_,datatype = "abundance", tree, q = c(0,1,2), reftime=NULL,
                       type = 'PD',level = NULL, nboot = 50,conf = 0.95)
```

The arguments of this function are described below.

<table style="width:100%;">
<colgroup>
<col width="20%">
<col width="80%">
</colgroup>
<thead>
<tr>
<th align="center">Argument</th>
<th align="left">Description</th>
</tr>
</thead>
<tbody>
<tr>
<td align="center"><code>data</code></td>
<td align="left"> a `matrix`/`data.frame` of species abundances or incidence data. See the description of function `iNEXTPD` for more details.</td>
</tr>
<tr>
<td align="center"><code>t_</code></td>
<td align="left"> needed only when `datatype = "incidence_raw"`, a sequence of named nonnegative integers specifying the sampling units in each community. Will be ignored if `datatype = "abundance"`.</td>
</tr>
<tr>
<td align="center"><code>datatype</code></td>
<td align="left"> the data type of input data: individual-based abundance data (`datatype = "abundance"`), or sampling-unit-based incidence matrix (`datatype = "incidence_raw"`). Default is `abundance`.</td>
</tr>
<tr>
<td align="center"><code>tree</code></td>
<td align="left"> a `phylo` object describing the Newick phylogeny tree for all observed species in the pooled assemblage.</td>
</tr>
<tr>
<td align="center"><code>q</code></td>
<td align="left"> a nonnegative value or sequence specifying the diversity order. Default is `c(0,1,2)`.</td>
</tr>
<tr>
<td align="center"><code>reftime</code></td>
<td align="left"> a positive value or sequence specifying the reference time for tree. If `NULL`, `reftime` = the tree depth of pooled assemblage. Default is `NULL`.</td>
</tr>
<tr>
<td align="center"><code>type</code></td>
<td align="left"> desired diversity type: `'PD'` for phylogenetic diversity and `'D'` for phylogenetic Hill number. See Chao et al. (2010) for details. Default is `'PD'`.</td>
</tr>
<tr>
<td align="center"><code>level</code></td>
<td align="left"> a positive value or sequence < 1 specifying a particular value of sample coverage. If `NULL`,then `level` will be chosen as the minimum coverage of all sites after extrapolating each site to its double sample sizes. Default is `NULL`.</td>
</tr>
<tr>
<td align="center"><code>nboot</code></td>
<td align="left"> a positive integer specifying the number of bootstrap replications. Enter 0 to skip bootstrap; in this case, the caculation of standard errors and confidence intervals will be skipped. Default is 50.</td>
</tr>
<tr>
<td align="center"><code>conf</code></td>
<td align="left"> a positive number < 1 specifying the level of confidence interval. Default is 0.95.</td>
</tr>


</tbody>
</table>

<br>


<br>


<br><font size="5"><b>PhD VIA EXAMPLES: Aundance data</b></font>
-------------------

Here, we use an abundance demo data as an illustrating example. If `datatype = "incidence_raw"`, slight modification of argument setting is needed; see the description of three main functions for detail. Some argument setting and output for incidence data also will be explained in later text.

Plants phylogeny and survey dataset is included in `PhD` package (`data(data.abu)`). The phylogeny tree (<code>\$tree</code>) of 46 plant species in this data set is constructed by the website phylomatic (http://phylodiversity.net/phylomatic/). This data includes three types of dunes habitats: emboryo dunes (ED), mobile dunes (MD) and transtion dunes (TD), which contains 17, 39, 42 distinct plants respectively. 
Throughout the illustration, we won't specify the argument `reftime` so that it is automatically fixed at 325, the tree depth of the pooled assemblage, by default.

To demonstrate this data, the following commands display basic data visualization:

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ade4)
library(ape)
library(dplyr)
```

```{r, fig.width=7, fig.height=7, fig.align="center"}
data(data.abu)
str(data.abu)
table.phylog(data.abu$data, ade4::newick2phylog(ape::write.tree(data.abu$tree)), 
             csize=4, f.phylog=0.7, clabel.row = .81)
```

Also, `PDInfo` tabulates some phylogenetic statistics for each assemblage, including reference sample size `n`, number of observed species `S.obs`, observed branch length (Faith's PD) `PD.obs`, the first two species frequency counts (`f1*`, `f2*`) and their branch length sums, respectively (`g1`, `g2`) at specified/default reference time `reftime`.

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth. 
PD_stat <- PDInfo(data = data.abu$data, tree = data.abu$tree, datatype = "abundance")
kable(PD_stat)
```

We now apply four main functions to the abundance data to see the outputs:

### 1. `iNEXTPD()`: 

```{r,}
dunes_iNEXT <- iNEXTPD(data = data.abu$data, tree = data.abu$tree, datatype = "abundance")
```

`iNEXTPD()` returns a tibble of PD or D estimates and sample coverage for interpolated or extrapolated sample sizes along with their confidence intervals (if `nboot > 0`). 

```{r,}
# only show part of the tibble, call View(dunes_iNEXT) to see full table.
kable(dunes_iNEXT[c(2,60,101,121,180,221,242,300,341),])
```

To visualize the curves, use `ggiNEXTPD` to obtain plots based on `ggplot2` package. This function provides three types of curves, which can be specified through the argument `plot.type`: 
1. Sample-size-based R/E curve (`plot.type = 1`): the curve of estimates as a function of sample size.
2. Coverage-based R/E curve (`plot.type = 2`): the curve of the estimates as a function of sample coverage.
3. Sample completeness curve (`plot.type = 3`): the curve of sample coverage with respect to sample size.}

Here, by default `plot.type = c(1,2,3)`.

```{r, eval= FALSE}
# plot.type = 1
ggiNEXTPD(dunes_iNEXT,plot.type = 1,datatype = 'abundance')$RE.plot.size
```

```{r,echo = FALSE ,fig.width=7, fig.height=7, fig.align="center"}
# plot.type = 1
ggiNEXTPD(dunes_iNEXT,plot.type = 1,datatype = 'abundance')$RE.plot.size + theme(text=element_text(size=14))
```

```{r, eval= FALSE}
# plot.type = 2
ggiNEXTPD(dunes_iNEXT,plot.type = 2,datatype = 'abundance')$RE.plot.C
```

```{r,echo = FALSE ,fig.width=7, fig.height=7, fig.align="center"}
ggiNEXTPD(dunes_iNEXT,plot.type = 2,datatype = 'abundance')$RE.plot.C + theme(text=element_text(size=14))
```

```{r, eval= FALSE}
# plot.type = 3
ggiNEXTPD(dunes_iNEXT,plot.type = 3,datatype = 'abundance')$RE.plot.sizeC
```

```{r,echo = FALSE ,fig.width=7, fig.height=7, fig.align="center"}
ggiNEXTPD(dunes_iNEXT,plot.type = 3,datatype = 'abundance')$RE.plot.sizeC + theme(text=element_text(size=14))
```


### 2. `PhdAsy()`

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth. 
dunes_Asy <- PhdAsy(data = data.abu$data, datatype = "abundance", 
                    tree = data.abu$tree, q = seq(0, 2, by = 0.25))
```

`PhdAsy()` function returns a tibble of estimated(asymptotic) phylogenetic diversity(`type = "PD"`) or phylogenetic Hill number(`type = "D"`)  with respect to specified/default order `q` and reference time `reftime`. It is based on statistical estimation of the unknown true PD or D; see Chao et al. (2015) and Hsieh and Chao (2017) for the statistical estimation detail.

```{r,}
# only show estimate for q = 0, 1, 2, call View(dunes_Asy) to see full table.
kable(dunes_Asy %>% filter(Order.q %in% c(0,1,2)))
```

To visualize the curves, use `ggtqplot` to obtain plots based on `ggplot2` package. This function provides two types of curves: order `q` profile or reference `time` profile, which can be specified through argument `profile`.

```{r, eval= FALSE}
ggtqplot(dunes_Asy,profile = 'q')
```

```{r,echo = FALSE, fig.width=7, fig.height=7, fig.align="center"}
ggtqplot(dunes_Asy,profile = 'q') + theme(text=element_text(size=14))
```

### 3. `PhdObs()`

For `PhdObs`, we demonstrate both observed order `q` profile and reference time profile of PD(`type = "PD"`). 

First, observed order `q` profile of PD with respect to tree depth, which is the default reference time point(`reftime`):

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth. 
dunes_Obs_q <- PhdObs(data = data.abu$data, datatype = "abundance", tree = data.abu$tree, 
                    q = seq(0, 2, by = 0.2),type = 'PD')
```

`PhdObs` function returns a tibble of empirical(observed) phylogenetic diversity(PD) or phylogenetic Hill number(D) with respect to specified/default order `q` and reference time `reftime`.

```{r,}
# only show estimate for q = 0, 1, 2, call View(dunes_Obs_q) to see full table.
kable(dunes_Obs_q %>% filter(Order.q %in% c(0,1,2)))
```

To visualize the curves, use `ggtqplot` to obtain plots based on `ggplot2` package. This function provides two types of curves: order `q` profile or reference `time` profile, which can be specified through argument `profile`.

```{r, eval= FALSE}
ggtqplot(dunes_Obs_q,profile = 'q')
```

```{r,echo = FALSE, fig.width=7, fig.height=7, fig.align="center"}
ggtqplot(dunes_Obs_q,profile = 'q') + theme(text=element_text(size=14))
```

Second, observed reference time profile of PD with respect to different order `q`:

```{r,}
dunes_Obs_time <- PhdObs(data = data.abu$data, datatype = "abundance", tree = data.abu$tree, 
                    q = c(0,1,2),reftime = seq(0.01,325,length.out = 40),type = 'PD')

# only show part of the tibble, call View(dunes_iNEXT) to see full table.
kable(head(dunes_Obs_time))
```

To visualize the curves, use `profile = 'time'` in `ggtqplot`.

```{r, eval= FALSE}
ggtqplot(dunes_Obs_time,profile = 'time')
```

```{r,echo = FALSE, fig.width=7, fig.height=7, fig.align="center"}
ggtqplot(dunes_Obs_time,profile = 'time') + theme(text=element_text(size=14))
```

### 4. `estimatePD()`

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth.
dunes_estPD <- estimatePD(data = data.abu$data, datatype = "abundance",
                          tree = data.abu$tree,q = c(0,1,2),level = 0.9)
```

The `estimatePD()` function returns a tibble of the PD or D estimates with respect to specified/default order `q` and reference time `reftime` for the user-specified sample coverages. In addition, the corresponding sample sizes and sample coverages are also provided.

```{r,}
kable(dunes_estPD)
```

<br><font size="5"><b>PhD VIA EXAMPLES: Incidence data</b></font>
-------------------

For incidence data (`datatype="incidence_raw"`), all four main functions in `PhD` can only accept incidence raw data (a species by sampling units matrix consisting of 0 and 1). 
Also, different from abundance data, users have to specify a sequence `t_` of N positve integer describing the number of sampling units for each of N assemblages. See the <b>DATA FORMAT/INFORMATION</b> section above for detailed required format.

Bird phylogeny and survey incidence dataset is included in `PhD` package (`data(data.inc)`). This data set describes the phylogeny (`$tree`) of 41 birds as reported by Jetz et al. (2012). It also gives the two sites (North.site and South.site) of species incidence (`$data`) data to these 41 species in November 2012 at Barrington Tops National Park, Australia. In `$data` matrix, the first 12 columns (sampling units) belongs to North.site and the left 17 columns (sampling units) belongs to South.site, which is described in `$t`.

Here, the procedures of applying four main functions to incidence data is similar to those for abundance case and their outcomes also differ slightly.

`PDInfo` tabulates some phylogenetic statistics for each assemblage, including number of sampling units `T`, number of observed species `S.obs`, observed branch length (Faith's PD) `PD.obs`, the first two species frequency counts (`Q1*`, `Q2*`) and their branch length sums, respectively (`R1`, `R2`) at specified/default reference time `reftime`.

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth. 
PD_stat <- PDInfo(data = data.inc$data, t_ = data.inc$t,datatype = "incidence_raw",tree = data.inc$tree)
kable(PD_stat)
```

We now apply four main functions to the incidence raw data to see the outputs:

### 1. `iNEXTPD()`: 

```{r,}
birds_iNEXT <- iNEXTPD(data = data.inc$data, t_ = data.inc$t,datatype = "incidence_raw",
                       tree = data.inc$tree)
```

`iNEXTPD()` returns a tibble of PD or D estimates and sample coverage for interpolated or extrapolated sample sizes along with their confidence intervals (if `nboot > 0`). 

```{r,}
# only show part of the tibble, call View(birds_iNEXT) to see full table.
kable(birds_iNEXT[c(6,36,61,78,123,158),])
```

To visualize the curves, use `ggiNEXTPD` to obtain plots based on `ggplot2` package. This function provides three types of curves, which can be specified through the argument `plot.type`: 
1. Sample-size-based R/E curve (`plot.type = 1`): the curve of estimates as a function of sample size.
2. Coverage-based R/E curve (`plot.type = 2`): the curve of the estimates as a function of sample coverage.
3. Sample completeness curve (`plot.type = 3`): the curve of sample coverage with respect to sample size.}

Here, by default `plot.type = c(1,2,3)`.

```{r, eval= FALSE}
# plot.type = 1
ggiNEXTPD(birds_iNEXT,plot.type = 1,datatype = 'incidence_raw')$RE.plot.size
```

```{r,echo = FALSE ,fig.width=7, fig.height=7, fig.align="center"}
# plot.type = 1
ggiNEXTPD(birds_iNEXT,plot.type = 1,datatype = 'incidence_raw')$RE.plot.size + theme(text=element_text(size=14))
```

```{r, eval= FALSE}
# plot.type = 2
ggiNEXTPD(birds_iNEXT,plot.type = 2,datatype = 'incidence_raw')$RE.plot.C
```

```{r,echo = FALSE ,fig.width=7, fig.height=7, fig.align="center"}
ggiNEXTPD(birds_iNEXT,plot.type = 2,datatype = 'incidence_raw')$RE.plot.C + theme(text=element_text(size=14))
```

```{r, eval= FALSE}
# plot.type = 3
ggiNEXTPD(birds_iNEXT,plot.type = 3,datatype = 'incidence_raw')$RE.plot.sizeC
```

```{r,echo = FALSE ,fig.width=7, fig.height=7, fig.align="center"}
ggiNEXTPD(birds_iNEXT,plot.type = 3,datatype = 'incidence_raw')$RE.plot.sizeC + theme(text=element_text(size=14))
```

### 2. `PhdAsy()`

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth. 
birds_Asy <- PhdAsy(data = data.inc$data, datatype = "incidence_raw", t_ = data.inc$t,
                    tree = data.inc$tree, q = seq(0, 2, by = 0.2))
```

`PhdAsy()` function returns a tibble of estimated(asymptotic) phylogenetic diversity(`type = "PD"`) or phylogenetic Hill number(`type = "D"`)  with respect to specified/default order `q` and reference time `reftime`. It is based on statistical estimation of the unknown true PD or D; see Chao et al. (2015) and Hsieh and Chao (2017) for the statistical estimation detail.

```{r,}
# only show estimate for q = 0, 1, 2, call View(birds_Asy) to see full table.
kable(birds_Asy %>% filter(Order.q %in% c(0,1,2)))
```

To visualize the curves, use `ggtqplot` to obtain plots based on `ggplot2` package. This function provides two types of curves: order `q` profile or reference `time` profile, which can be specified through argument `profile`.

```{r, eval= FALSE}
ggtqplot(birds_Asy,profile = 'q')
```

```{r,echo = FALSE, fig.width=7, fig.height=7, fig.align="center"}
ggtqplot(birds_Asy,profile = 'q') + theme(text=element_text(size=14))
```

### 3. `PhdObs()`

For `PhdObs`, we demonstrate both observed order `q` profile and reference time profile of PD(`type = "PD"`). 

First, observed order `q` profile of PD with respect to tree depth, which is the default reference time point(`reftime`):

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth. 
birds_Obs_q <- PhdObs(data = data.inc$data, t_ = data.inc$t ,datatype = "incidence_raw",
                      tree = data.inc$tree, q = seq(0, 2, by = 0.2),type = 'PD')
```

`PhdObs` function returns a tibble of empirical(observed) phylogenetic diversity(PD) or phylogenetic Hill number(D) with respect to specified/default order `q` and reference time `reftime`.

```{r,}
# only show estimate for q = 0, 1, 2, call View(birds_Obs_q) to see full table.
kable(birds_Obs_q %>% filter(Order.q %in% c(0,1,2)))
```

To visualize the curves, use `ggtqplot` to obtain plots based on `ggplot2` package. This function provides two types of curves: order `q` profile or reference `time` profile, which can be specified through argument `profile`.

```{r, eval= FALSE}
ggtqplot(birds_Obs_q,profile = 'q')
```

```{r,echo = FALSE, fig.width=7, fig.height=7, fig.align="center"}
ggtqplot(birds_Obs_q,profile = 'q') + theme(text=element_text(size=14))
```

Second, observed reference time profile of PD with respect to different order `q`:

```{r,}
birds_Obs_q <- PhdObs(data = data.inc$data,t_ = data.inc$t, datatype = "incidence_raw", 
                         tree = data.inc$tree,q = c(0,1,2),
                         reftime = seq(0.01,82.8575,length.out = 40),type = 'PD')

# only show part of the tibble, call View(birds_Obs_q) to see full table.
kable(head(birds_Obs_q))
```

To visualize the curves, use `profile = 'time'` in `ggtqplot`.

```{r, eval= FALSE}
ggtqplot(birds_Obs_q,profile = 'time')
```

```{r,echo = FALSE, fig.width=7, fig.height=7, fig.align="center"}
ggtqplot(birds_Obs_q,profile = 'time') + theme(text=element_text(size=14))
```

### 4. `estimatePD()`

```{r,}
# leave argument 'reftime' unsecified and it will be defaulted to the tree depth.
birds_estPD <- estimatePD(data = data.inc$data,t_ = data.inc$t,datatype = "incidence_raw",
                          tree = data.inc$tree,q = c(0,1,2),level = 0.9)
```

The `estimatePD()` function returns a tibble of the PD or D estimates with respect to specified/default order `q` and reference time `reftime` for the user-specified sample coverages. In addition, the corresponding sample sizes and sample coverages are also provided.

```{r,}
# only show part of the tibble, call View(birds_estPD) to see full table.
kable(birds_estPD)
```


<br><font size="4"><b>License</b></font>
-------------------
The PhD package is licensed under the GPLv3. To help refine `PhD`, your comments or feedbacks would be welcome (please send them to Anne Chao).

<br><font size="4"><b>How to cite</b></font>
-------------------
If you publish your work based on results from `PhD` (R package), please make reference to Chao et al. given in the following list.

<br><font size="4"><b>Reference</b></font>
-------------------
- Chao, A., Chiu C.-H. and Jost, L. (2010). Phylogenetic diversity measures based on Hill numbers. Philosophical Transactions of the Royal Society B., 365, 3599-3609.
- Chiu, C.-H., Jost, L. and Chao, A. (2014). Phylogenetic beta diversity, similarity, and differentiation measures based on Hill numbers. Ecological Monographs, 84, 21-44.
- Chao, A., Chiu, C.-H., Hsieh, T. C., Davis, T., Nipperess, D., and Faith, D. (2015) Rarefaction and extrapolation of phylogenetic diversity. Methods in Ecology and Evolution, 6, 380-388.
- Hsieh, T. C. and Chao, A. (2017). Rarefaction and extrapolation: making fair comparison of abundance-sensitive phylogenetic diversity among multiple assemblages. Systematic Biology 66, 100-111.

