---
title: "A Quick Introduction to PhD via Examples"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{A Quick Introduction to PhD via Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE, message = FALSE,
  comment = "#>"
)
library(PhDyhc)
library(ggplot2)
data(data.abu)
data(data.inc)
```
`PhD`(Phylogenetic Diversity) is a R package providing functions for rarefaction/extrapolation and asymptotic estimation of phylogenetic diversity.(+ also emprircal, phylogenetic Hill numbers) 3 main functions are provided:

1. `inextPD`: the phylogenetic version of `iNEXT`, focusing on the seamless rarefaction and extrapolation sampling curves of phylogenetic diversity for q = 0, 1 and 2. See Chao et al. (2010, 2015) and Hsieh and Chao (2017) for pertinent background and methods.

2. `PhdAsy`: provides the estimated asymptotic phylogenetic q profile. 
3. `PhdObs`: provides the empirical q and time profix


```{r setup}
library(PhD)
```
